Emittance preservation of an electron beam in a loaded quasi-linear plasma wakefield

### 2017-10-05 Tex/Paper.tex
More cleanup, reflow, and figure text updates

% Text
%~ \newcommand{\eq}[1]{Eq. \ref{#1}}
%~ \newcommand{\fig}[1]{Fig. \ref{#1}}
%~ \newcommand{\tbl}[1]{Table \ref{#1}}

@@

%Patric general comments
%energy labels not consistent between figs 7, 9 and 8
%labels for the fractional charge is too complicated on Figs 7,8,9. it needs to be explained in the caption and/or in the text. "On the left axis we plot the fraction of that charge whose emittance remains .... "
%Add scale for grey color on Fig. 2
%Fig. 3 has a "v=c" and an arrow that are only on that figure ...
%It is not clear to me what the conclusions/outcome of the last section are ....

@@

\begin{abstract}
We investigate beam loading and emittance preservation for a high-charge electron beam being accelerated in quasi-linear
plasma wakefield driven by a short proton beam. The structure of the studied wakefields are similar to those of a long,
modulated proton beam, such as the AWAKE proton driver. We show that by properly choosing the electron beam parameters
and exploiting two well known effects, beam loading of the wake field and full blow out of plasma electrons by the
accelerated beam, the electron beam can %PM be accelerated - EA ok
gain large amounts of energy with a narrow final energy spread (\%-level) and without significant emittance growth.
\end{abstract}

@@

Beam driven plasma wakefield accelerators \cite{chen:1985} have the potential to offer compact linear accelerators with
high energy gradients, and have been of interest for several decades. With a relativistic charged particle drive
beam travelling through a plasma, a strong wakefield is excited, that can be loaded by a trailing witness beam. When the
witness beam optimally loads the wakefield, an increase in absolute energy spread can be kept small. %to a minimum. %PM - EA ok
The concept has been demonstrated experimentally in the past using electron drive beams accelerating electron witness
beams \cite{rosenzweig:1988, blumenfeld:2007, kallos:2008, litos:2014}.

A major challenge with plasma wakefield accelerators is, however, to accelerate a beam while keeping both energy spread
and emittance growth small. %growth to a minimum. PM - EA ok
In the well described linear regime, valid when the beam density $n_{b}$ is much smaller than the plasma density
$n_{0}$, a non-linear transverse focusing force causes emittance growth of the witness beam. The beam
%will %PM - EA ok
also sees a transversely and longitudinally varying accelerating field causing a spread in energy after the beam has
been accelerated \cite{katsouleas:1987}. In the non-linear regime, where $n_{b} > n_{0}$, a bubble is formed by the
transverse oscillations of the plasma electrons, %forming PM - EA ok
gathering in a sheet around an evacuated area filled with only ions. The ions, assumed stationary, form an uniform
density ion channel creating a focusing force that varies linearly with radius. This %produces a  PM - EA ok
focusing force %that PM - EA ok
preserves emittance \cite{rosenzweig:1991}.
% EA: I put Rosenzweig original blow out paper as citation here instead of, \cite{lu:2006-1, lu:2006}.

In this paper we present simulation results showing how emittance preservation of a high charge density witness beam can
be ensured, when accelerated by a proton drive beam producing quasi-linear wakefields \cite{rosenzweig:2010}. By
quasi-linear wakefields we here mean wakefields with only partial blow out of the plasma electrons in the bubble.
%~ Patric's version is implemented:
The key idea is to have enough charge in the witness beam to at the same time load the wakefield to produce low
relative energy spread%$\Delta E/E$, 
, and completely blow out the electrons left in the bubble after the beam to reach conditions that preserves emittance.
The results have importance for the preparation of AWAKE Run 2 and possibly other applications in the quasi-linear regime.

@@

The energy carried by electron drive bunches used in previous plasma wakefield experiments have been no more than
$100\unit{J}$ and the propagation length typically no more than $1\unit{m}$ \cite{blumenfeld:2007,caldwell:2009}. For
high-energy physics application a higher total beam energy is often desired, for instance, the energy of a high-charge
electron bunch accelerated to $1\unit{TeV}$ with $1\nexp{10}$ electrons, similar to the beam that could be produced by
the international linear collider, is $1.6\unit{kJ}$. Using electron beams as drivers a large number of plasma stages is
required to reach %PM - EA ok
an energy of a $\unit{kJ}$ for the accelerated bunch. However, staging plasma accelerators without reducing the
effective gradient and spoiling the beam quality is challenging \cite{steinke:2016,lindstrom:2016}.

Proton beams available at CERN carry significantly more energy than available electron beams, $19\unit{kJ}$ for the SPS
beam \cite{gschwendtner:2016}, allowing for much longer plasma wakefield accelerator stages. The SPS beam is orders of
magnitude longer than the plasma wavelengths needed for such applications, and the beam %will PM - EA ok
does not drive a strong wake \cite{gschwendtner:2016}.  By letting the proton beam undergo self-modulation before
injecting the witness beam into accelerating structure, %most PM - EA ok
stronger wakefields are %formed PM - EA ok
excited. The self-modulation of the beam is produced by the transverse fields generated by the beam acting upon itself,
causing regions of the beam to rapidly defocus \cite{kumar:2010}. The modulation frequency is close to that of the
plasma, and produces a train of short proton bunches along the beam axis with a surrounding halo of defocused particles.
This train of bunches resonantly %produces - EA ok
drives wakefields to large %r PM - EA ok 
amplitudes.

AWAKE at CERN is a proof of concept proton beam driven plasma wakefield accelerator experiment
\cite{awake_collaboration:2014}, currently in its first phase of operation. The experiment uses a $400\unit{GeV}$ proton
beam delivered by the SPS as its driver, and a single $10\unit{m}$ plasma stage with a nominal plasma density of
$7\nexp{14}\unit{cm}^{-3}$ \cite{gschwendtner:2016}. This plasma density corresponds to $\lambda_{pe} = 1.26\unit{mm}$
-- matched to the transverse size of the SPS proton beam such that $k_{pe}\sigma_{x,y,pb} = 1$, where
$k_{pe} = 2\pi/\lambda_{pe}$ is the plasma wave number, %PM. - EA ok
$\lambda_{pe}=2\pi c/\omega_{pe}$ and $\omega_{pe}=\left(n_0e^2/m_e\epsilon_0\right)^{1/2}$ is the plasma electron
angular frequency.

@@

The focus in this study is the beam loading of the wakefields driven by the proton beam. Studies of self-modulated
proton bunches show that the proton bunch evolve as it propagates through a uniform plasma \cite{lotov:2011}, but that
small variations in the plasma profile the modulation, and thus the wake fields, may be stabilized over long distances
\cite{lotov:2011, lotov:2015, caldwell:2011}.  To study the witness beam evolution in a stable wake, independent of the
dynamics of the self-modulation, %PMwe define here - EA ok
here we use a single, non-evolving proton bunch as driver. The proton bunch parameters are chosen so that key features of
the wake -- the plasma electron density in the wake and the longitudinal electric field -- are the same as %PM - EA ok
in the wake of a self-modulated proton beam with AWAKE baseline parameters \cite{gschwendtner:2016}. Both the proton
beam and the witness have Gaussian %PM - EA ok
longitudinal charge distribution and bi-Gaussian transverse charge distributions.

We have previously studied the beam loading in a proton beam wake using the full PIC code Osiris \cite{fonseca:2002}
%using PM - EA ok
and with 2D cylindrical-symmetric simulations. The studies \cite{berglyd_olsen:2015, berglyd_olsen:2016} primarily
looked at beam loading, energy gain and energy spread, as well as different approaches to create a stable drive beam
structure based on previous self-modulation studies. In order to study the witness beam emittance evolution we adopt the
recently released open source version of QuickPIC \cite{huang:2006, an:2013}. QuickPIC is a fully relativistic 3D
quasi-static PIC code. It does not suffer from the numerical Cherenkov effect that full PIC codes do
\cite{godfrey:1974,lehe:2013}, making it a well suited tool to study emittance preservation.  All simulation results in
this paper are %PMcalculated - EA ok
obtained using QuickPIC open source.

@@

%~ In these simulations we use a single proton drive beam that sets up a wakefield comparable to that which we expect to
%~ see from the self-modulated SPS beam in the second plasma cell of AWAKE Run 2 (Figure \ref{Fig:AWAKER2}).

%PM The baseline AWAKE drive beam current is insufficient to reach the non-linear regime and produce a plasma bubble. - EA ok
The modulation process used in AWAKE does not reach the fully non-linear regime and thus does not produce a bubble void
of plasma electrons. When the SPS beam, containing $3\nexp{11}$ protons \cite{gschwendtner:2016}, enters the second
plasma stage (figure \ref{Fig:AWAKER2}), the peak electric field is expected to be about $500\funit{MV}{m}$. The plasma
electron density are only depleted to around $65\%$ of nominal plasma density at the point where we inject the electron
beam \cite{awake_collaboration:2016}.  The plasma electron density depletion and the peak field is replicated using a
single %PM proton 
bunch with $1.46\nexp{10}$ protons (or $2.34\unit{nC}$), a %PM bunch
length $\sigma_{z} = 40\unit{\mu m}$ ($7\unit{kA}$), and a transverse size $\sigma_{x,y,pb} = 200\unit{\mu m}$.
The peak density is $0.83\cdot n_{0}$ results in a quasi-linear wake. %PM The beam energy is $400\unit{GeV}$.
To avoid transverse evolution of the proton driver, emulating the stable propagation of the self-modulated proton beam
\cite{lotov:2011, lotov:2015, caldwell:2011} we freeze the transverse evolution of the equivalent proton bunch by
increasing the particles mass by six orders of magnitude.

@@

%The small size of the accelerating structure produced by the self-modulated beam also
%puts constraints on the size of the electron witness beam, and thus its charge and current, if at the same time we wish
%to prevent large energy spread.  
%PM not sure what the point of the above sentence is ... - EA ok
In order to prevent large amplitude oscillations of the witness beam particles, which may cause additional energy spread
as well as emittance growth, we %PMstudy
consider a witness beam matched to the plasma density.  
%The beam matching relation is given by
%\begin{equation}
%    \beta = \frac{\sigma_{x,y}^{2}\gamma}{\epsilon_{N}}
%          = \frac{c}{\omega_{pe}}\sqrt{2\gamma}, \label{EQ:MatchedB}
%    % = c\sqrt{\frac{m_{e}\varepsilon_{0}}{n_{pe}e^{2}}2\gamma}
%\end{equation}
%where $\beta$ is the Twiss parameter. It then follows that the transverse profile of the beam must satisfy
%PM Not sure what the point of the above lines is ....
The matched beam transverse size is:
\begin{equation}
    \sigma_{x,y,eb}=\left(\frac{2c^{2}\epsilon_{\mathrm{N}}^{2}m_{e}\varepsilon_{0}}{n_{pe}e^{2}\gamma}\right)^{1/4}
    \label{EQ:Matched}
\end{equation}

We %furthermore PM - EA ok
assume an initial normalized emittance of $\epsilon_{\mathrm{N}} = 2\unit{\mu m}$. This emittance is possible to produce
with a standard RF-injector, while at the same time yielding a sufficient narrow beam. % EA: comment, perhaps put a ref here [***ref].

Beam loading of a short witness beam is sensitive to its position relative to the electric field \cite{tzoufras:2009}
as well as to its dephasing with respect to the wakefields when at low energy.
%PM A low gamma witness beam can easily drift out of phase with the drive beam wakefield. The drift of the beams is proportional to $\gamma^{-2}$,  
To eliminate %initial 
de-phasing of the witness beam, % PM
the initial beam energy %PMwas
is set such
that $\gamma_{eb} = \gamma_{pb} = 426.3$, giving an energy of $217\unit{MeV}$.  A lower initial energy is likely to be
sufficient for AWAKE Run 2 injection. 
% EA: we have studied this to some extent, but unfortunately we do not have a good refence for the last statement, no details in \cite{berglyd_olsen:2015}.

Eq. (\ref{EQ:Matched}) yields a transverse size of $\sigma_{x,y,eb}$ of $5.25\unit{\mu m}$, narrow compared to the drive
beam $\sigma_{x,y,pb} = 200\unit{\mu m}$. The bunch length was set to $\sigma_{z}=60\unit{\mu m}$, based on earlier
beam loading studies \cite{berglyd_olsen:2016}. The charge %PM was
is adjusted to $100\unit{pC}$ for optimal beam loading, as discussed in the next section. We refer to the defined drive
beam and witness beam parameter set as the \emph{base case}.  
% EA: the key figure, figure 2, was not discussed in the text until very far down!  I therefore added (text could possibly be improved)
Figure \ref{Fig:PlasmaDenTWake} shows the two beams from QuickPIC simulations of the initial time step, for the base
case parameters; the proton bunch in blue, the trailing electron beam in red and the plasma electron density in grey.

@@

The relatively small size of the witness beam %PM compared to the proton beam put some restrictions on the transverse resolution of the witness beam as we need a simulation box wide enough to contain the larger beam while resolving the narrower one (see figure \ref{Fig:PlasmaDenTWake}). 
puts constraints on the transverse cell size and number in the simulations: small size to resolve the small electron
beam and large number of to resolve the large proton beam and the wakefields.
We use %PMd  - EA ok
a transverse grid cell size of $1.17\unit{\mu m}$, and of $2.34\unit{\mu m}$ for the longitudinal grid cells for the
simulations presented in section \ref{S:BL}. The witness beam was simulated with $16.8\nexp{6}$ non-weighted particles.
%  ***NN*** plasma particles were simulated per grid cell.
% EA: Veronica to fill in NN above
% VO: QuickPIC doesn't specify particles per cell. Only in Osiris,

\begin{figure}[hbt]
    \includegraphics[width=\linewidth,trim={2mm 0mm 2mm 0mm},clip]{figures/beamLoading}
    \caption{\label{Fig:BeamLoading} Top plot%PM plot shows the - EA ok
    : unloaded longitudinal electric field with no witness beam (dashed blue line) and %PMthe
        loaded field (whole blue line) along the beam axis. The beam density along the axis for both beams are shown in
        red. %PM The bottom plot compares the 
        Bottom plot: plasma densities along the beam axis for a drive beam with no witness beam (dashed green line),
        witness beam with no drive beam (dash-dotted green line), and both beams present (whole green line).
        $\xi = z - tc$ is the position in the simulation box, moving towards the left.}
\end{figure}

\begin{figure}[hbt]
    \includegraphics[width=\linewidth,trim={2mm 0mm 2mm 0mm},clip]{figures/beamPhaseSpace}
    \caption{\label{Fig:BeamPS} %PMThe p
        Phase space charge distribution of a $100\unit{pC}$, $60\unit{\mu m}$ long witness beam after $4\unit{m}$ of
        plasma. The mean momentum is $1.67\unit{GeV/c}$ with an RMS energy spread of $87\unit{MeV/c}$ ($5.2\%$) for the
        full beam.}
\end{figure}

@@

\begin{figure*}[hbt]
    \includegraphics[width=\linewidth,trim={0mm 0mm 0mm 0mm},clip]{figures/beamFilamentationAll}
    \caption{\label{Fig:BeamFilament} Plots \textbf{a} to \textbf{f} show the transverse phase space of the offset
        electron beam at different plasma positions. Plot \textbf{g} shows the macro particle mean position, and plot
        \textbf{h} their RMS spread. Plots \textbf{a}, \textbf{b} and \textbf{c}, as well as the blue lines in plots
        \textbf{g} and \textbf{h} represent particles %PM
        not in the ion column (see Fig.~\ref{Fig:BeamEmitt}), with position $1.40\unit{\mu m} < \xi < 1.42\unit{\mu m}$.
        Plots \textbf{d}, \textbf{e} and \textbf{f}, and the red lines in plots \textbf{g} and \textbf{h} represent
        particles in the ion column with position $1.55\unit{\mu m} < \xi < 1.57\unit{\mu m}$.}
\end{figure*}

@@

Figure \ref{Fig:BeamLoading} shows the results of QuickPIC simulations of the initial time step, for the base case
parameters. The $E_{z}$-field generated by the proton drive beam is seen as the blue line, shown with and without the
electron beam present. With a proton beam density $n_{pb} \simeq n_{0}$, %PMwe 
the wakefields are in the quasi-linear regime \cite{rosenzweig:2010}. The dashed green line in the lower part of figure
\ref{Fig:BeamLoading} shows that the on-axis plasma density has a depletion of $67\%$, close to what we see in full
scale reference simulations for AWAKE Run 2 \cite{awake_collaboration:2016}.

@@

The witness beam generates its own wakefield which %PM partially cancels out 
loads the $E_{z}$ field generated by the drive beam. With an ideally shaped electron beam charge profile it is possible
to optimally load the field in such a way that the accelerating field is constant along the beam
\cite{katsouleas:1987, tzoufras:2009}. Gaussian beams, as assumed in these studies, cannot completely flatten the
electric field in the tails of the charge distribution, and our base case beam therefore has a tail in energy both at
the front and the back of the beam, as illustrated in figure \ref{Fig:BeamPS}. The bulk of the beam, however, sees a
relatively flat field.
% EA: I rewrote this to the sentence after the comments
%we can acceleratoreHowever, with such a narrow beam when matched beam to the plasma density (equation
%\ref{EQ:MatchedB}), we are limited on how much total charge we can accelerate without reaching charge densities that
%will overload the field.
%  EA: shall we say the below?  The sentence asks for more detailed studies of the dependence of the loading of the transverse fields.
%The amounts of charge required to load the field depends both on the witness bunch length as well as the transverse size.

The %PM corresponding 
electron beam density is $n_{eb} \approx 35\times n_{0}$. This means that the witness beam's own wakefield is in the
fully non-linear regime, where the space charge force is sufficient to blow-out all plasma electrons resulting in the
formation of an ion column %PM 
(Fig.~\ref{Fig:BeamLoading} bottom). This ion column, as is well known \cite{rosenzweig:1991}, provides a linear
focusing force on the part the electron beam within the column, and therefore prevents emittance growth for this %PMregion
part of the beam.  This effect is shown for our base case in figure \ref{Fig:PlasmaDenTWake}. The focusing field has a
gradient of $20\unit{kT/m}$ near the beam axis, corresponding to the matched field gradient.

@@

%~ The witness beam's self-generated bubble prevents the tailing end of the beam to see any significant emittance growth.
% EA: I cut the line below, we are still discussing the base case
%For the majority of the cases we studied, that maintained a stable accelerating structure, 
 %PMDue to the ion column focusing, 
Simulation results show that about $70-80\%$ of the electron beam longitudinal slices retain their initial emittance
after the propagation in the plasma. Figure \ref{Fig:BeamEmitt} shows the slice emittance along the beam for the base
case, sampled after propagating through $0$, $4$, $40$ and $100\unit{m}$ of plasma. Emittance growth mainly occurs in
first few meters, and no significant emittance growth %PMwas
is observed after this for propagation lengths up to $100\unit{m}$. The head of the beam does not benefit from the full
ion column focusing, but since the proton beam creates a quasi-linear wake, the emittance of the head of the beam still
stabilizes after some time. For the 100 meter simulation, the drive beam energy was increased to $7\unit{TeV}$
(LHC energy) to prevent de-phasing, as de-phasing starts to become a significant effect for the SPS beam of
$400\unit{GeV}$ after about $50\unit{m}$.

>>>>

@@

%Patric general comments
%  energy labels not consistent between figs 7, 9 and 8
%  labels for the fractional charge is too complicated on Figs 7,8,9. it needs to be explained in the caption and/or in
%  the text. "On the left axis we plot the fraction of that charge whose emittance remains .... "
%  Add scale for grey color on Fig. 2
%  Fig. 3 has a "v=c" and an arrow that are only on that figure ...
%  It is not clear to me what the conclusions/outcome of the last section are ....

@@

\begin{abstract}
We investigate beam loading and emittance preservation for a high-charge electron beam being accelerated in quasi-linear
plasma wakefield driven by a short proton beam. The structure of the studied wakefields are similar to those of a long,
modulated proton beam, such as the AWAKE proton driver. We show that by properly choosing the electron beam parameters
and exploiting two well known effects, beam loading of the wake field and full blow out of plasma electrons by the
accelerated beam, the electron beam can gain large amounts of energy with a narrow final energy spread (\%-level) and
without significant emittance growth.
\end{abstract}

@@

Beam driven plasma wakefield accelerators \cite{chen:1985} have the potential to offer compact linear accelerators with
high energy gradients, and have been of interest for several decades. With a relativistic charged particle drive beam
travelling through a plasma, a strong wakefield is excited, that can be loaded by a trailing witness beam. When the
witness beam optimally loads the wakefield, an increase in absolute energy spread can be kept small. The concept has
been demonstrated experimentally in the past using electron drive beams accelerating electron witness beams
\cite{rosenzweig:1988, blumenfeld:2007, kallos:2008, litos:2014}.

A major challenge with plasma wakefield accelerators is, however, to accelerate a beam while keeping both energy spread
and emittance growth small. In the well described linear regime, valid when the beam density $n_{b}$ is much smaller
than the plasma density $n_{0}$, a non-linear transverse focusing force causes emittance growth of the witness beam.
The beam also sees a transversely and longitudinally varying accelerating field causing a spread in energy after the
beam has been accelerated \cite{katsouleas:1987}. In the non-linear regime, where $n_{b} > n_{0}$, a bubble is formed by
the transverse oscillations of the plasma electrons, gathering in a sheet around an evacuated area filled with only
ions. The ions, assumed stationary, form an uniform density ion channel creating a focusing force that varies linearly
with radius. This focusing force preserves emittance \cite{rosenzweig:1991}.

In this paper we present simulation results showing how emittance preservation of a high charge density witness beam can
be ensured, when accelerated by a proton drive beam producing quasi-linear wakefields \cite{rosenzweig:2010}. By
quasi-linear wakefields we here mean wakefields with only partial blow out of the plasma electrons in the bubble.
The key idea is to have enough charge in the witness beam to at the same time load the wakefield to produce low relative
energy spread, and completely blow out the electrons left in the bubble after the beam to reach conditions that
preserves emittance. The results have importance for the preparation of AWAKE Run 2 and possibly other applications in
the quasi-linear regime.

@@

The energy carried by electron drive bunches used in previous plasma wakefield experiments have been no more than
$100\unit{J}$ and the propagation length typically no more than $1\unit{m}$ \cite{blumenfeld:2007,caldwell:2009}.
% VKBO: Not sure "no more than" is true.
For high-energy physics application a higher total beam energy is often desired, for instance, the energy of a
high-charge electron bunch accelerated to $1\unit{TeV}$ with $1\nexp{10}$ electrons, similar to the beam that could be
produced by the international linear collider, is $1.6\unit{kJ}$. Using electron beams as drivers a large number of
plasma stages is required to reach an energy of a $\unit{kJ}$ for the accelerated bunch. However, staging plasma
accelerators without reducing the effective gradient and spoiling the beam quality is challenging
\cite{steinke:2016,lindstrom:2016}.

Proton beams available at CERN carry significantly more energy than available electron beams, $19\unit{kJ}$ for the SPS
beam \cite{gschwendtner:2016}, allowing for much longer plasma wakefield accelerator stages. The SPS beam is orders of
magnitude longer than the plasma wavelengths needed for such applications, and the beam does not drive a strong wake
\cite{gschwendtner:2016}. By letting the proton beam undergo self-modulation before injecting the witness beam into the % added "the"
accelerating structure, stronger wakefields are excited. The self-modulation of the beam is produced by the transverse
fields generated by the beam acting upon itself, causing regions of the beam to rapidly defocus \cite{kumar:2010}. The
modulation frequency is close to that of the plasma, and produces a train of short proton bunches along the beam axis
with a surrounding halo of defocused particles. This train of bunches resonantly drives wakefields to large amplitudes.

AWAKE at CERN is a proof of concept proton beam driven plasma wakefield accelerator experiment
\cite{awake_collaboration:2014}, currently in its first phase of operation. The experiment uses a $400\unit{GeV}$ proton
beam delivered by the SPS as its driver, and a single $10\unit{m}$ plasma stage with a nominal plasma density of
$7\nexp{14}\unit{cm}^{-3}$ \cite{gschwendtner:2016}. This plasma density corresponds to $\lambda_{pe} = 1.26\unit{mm}$
-- matched to the transverse size of the SPS proton beam such that $k_{pe}\sigma_{x,y,pb} = 1$, where
$k_{pe} = 2\pi/\lambda_{pe}$ is the plasma wave number, $\lambda_{pe}=2\pi c/\omega_{pe}$ and
$\omega_{pe}=\left(n_0e^2/m_e\epsilon_0\right)^{1/2}$ is the plasma electron angular frequency.

@@

The focus in this study is the beam loading of the wakefields driven by the proton beam. Studies of self-modulated
proton bunches show that the proton bunch evolve as it propagates through a uniform plasma \cite{lotov:2011}, but that
small variations in the plasma profile the modulation, and thus the wake fields, may be stabilised over long distances
\cite{lotov:2011, lotov:2015, caldwell:2011}. To study the witness beam evolution in a stable wake, independent of the
dynamics of the self-modulation, here we use a single, non-evolving proton bunch as driver. The proton bunch parameters
are chosen so that key features of the wake -- the plasma electron density in the wake and the longitudinal electric
field -- are the same as in the wake of a self-modulated proton beam with AWAKE baseline parameters
\cite{gschwendtner:2016}. Both the proton beam and the witness have Gaussian longitudinal charge distribution and
bi-Gaussian transverse charge distributions.

We have previously studied the beam loading in a proton beam wake using the full PIC code Osiris \cite{fonseca:2002}
and with 2D cylindrical-symmetric simulations. The studies \cite{berglyd_olsen:2015, berglyd_olsen:2016} primarily
looked at beam loading, energy gain and energy spread, as well as different approaches to create a stable drive beam
structure based on previous self-modulation studies. In order to study the witness beam emittance evolution we adopt the
recently released open source version of QuickPIC \cite{huang:2006, an:2013}. QuickPIC is a fully relativistic 3D
quasi-static PIC code. It does not suffer from the numerical Cherenkov effect that full PIC codes do
\cite{godfrey:1974,lehe:2013}, making it a well suited tool to study emittance preservation.  All simulation results in
this paper are obtained using QuickPIC open source.

@@

%~ In these simulations we use a single proton drive beam that sets up a wakefield comparable to that which we expect to
%~ see from the self-modulated SPS beam in the second plasma cell of AWAKE Run 2 (Figure \ref{Fig:AWAKER2}).
%PM The baseline AWAKE drive beam current is insufficient to reach the non-linear regime and produce a plasma bubble.

The modulation process used in AWAKE does not reach the fully non-linear regime and thus does not produce a bubble void
of plasma electrons. When the SPS beam, containing $3\nexp{11}$ protons \cite{gschwendtner:2016}, enters the second
plasma stage (figure \ref{Fig:AWAKER2}), the peak electric field is expected to be about $500\funit{MV}{m}$. The plasma
electron density are only depleted to around $65\%$ of nominal plasma density at the point where we inject the electron
beam \cite{awake_collaboration:2016}.  The plasma electron density depletion and the peak field is replicated using a
single bunch with $1.46\nexp{10}$ protons (or $2.34\unit{nC}$), a length $\sigma_{z} = 40\unit{\mu m}$ ($7\unit{kA}$),
and a transverse size $\sigma_{x,y,pb} = 200\unit{\mu m}$. The peak density is $0.83\cdot n_{0}$ results in a
quasi-linear wake. To avoid transverse evolution of the proton driver, emulating the stable propagation of the
self-modulated proton beam \cite{lotov:2011, lotov:2015, caldwell:2011} we freeze the transverse evolution of the
equivalent proton bunch by increasing the particles mass by six orders of magnitude.

@@

In order to prevent large amplitude oscillations of the witness beam particles, which may cause additional energy spread
as well as emittance growth, we consider a witness beam matched to the plasma density. The matched beam transverse size
is:
\begin{equation}
    \sigma_{x,y,eb}=\left(\frac{2c^{2}\epsilon_{\mathrm{N}}^{2}m_{e}\varepsilon_{0}}{n_{pe}e^{2}\gamma}\right)^{1/4}
    \label{EQ:Matched}
\end{equation}

We assume an initial normalized emittance of $\epsilon_{\mathrm{N}} = 2\unit{\mu m}$. This emittance is possible to
produce with a standard RF-injector, while at the same time yielding a sufficient narrow beam.
% EA: comment, perhaps put a ref here [***ref].

Beam loading of a short witness beam is sensitive to its position relative to the electric field \cite{tzoufras:2009}
as well as to its de-phasing with respect to the wakefields when at low energy. To eliminate de-phasing of the witness
beam, the initial beam energy is set such that $\gamma_{eb} = \gamma_{pb} = 426.3$, giving an energy of $217\unit{MeV}$.
A lower initial energy is likely to be sufficient for AWAKE Run 2 injection.
% EA: we have studied this to some extent, but unfortunately we do not have a good refence for the last statement,
%     no details in \cite{berglyd_olsen:2015}.

Eq. (\ref{EQ:Matched}) yields a transverse size of $\sigma_{x,y,eb}$ of $5.25\unit{\mu m}$, narrow compared to the drive
beam $\sigma_{x,y,pb} = 200\unit{\mu m}$. The bunch length was set to $\sigma_{z}=60\unit{\mu m}$, based on earlier beam
loading studies \cite{berglyd_olsen:2016}. The charge is adjusted to $100\unit{pC}$ for optimal beam loading, as
discussed in the next section. We refer to the defined drive beam and witness beam parameter set as the
\emph{base case}.
% EA: the key figure, figure 2, was not discussed in the text until very far down!
%     I therefore added (text could possibly be improved)
Figure \ref{Fig:PlasmaDenTWake} shows the two beams from QuickPIC simulations of the initial time step, for the base
case parameters; the proton bunch in blue, the trailing electron beam in red and the plasma electron density in grey.

@@

The relatively small size of the witness beam puts constraints on the transverse cell size and number in the
simulations: small size to resolve the small electron beam and large number of to resolve the large proton beam and the
wakefields. % This sounds really bad? Should be improved.
We use a transverse grid cell size of $1.17\unit{\mu m}$, and of $2.34\unit{\mu m}$ for the longitudinal grid cells for
the simulations presented in section \ref{S:BL}. The witness beam was simulated with $16.8\nexp{6}$ non-weighted
particles.
%  ***NN*** plasma particles were simulated per grid cell.
% EA: Veronica to fill in NN above
% VO: QuickPIC doesn't specify particles per cell. Only in Osiris, This number has no fixed value, just a range from
%     zero to whatever is the density at the peak of the Gaussian distribution.

\begin{figure}[hbt]
    \includegraphics[width=\linewidth,trim={2mm 0mm 2mm 0mm},clip]{figures/beamLoading}
    \caption{\label{Fig:BeamLoading} \textbf{Top plot:} unloaded longitudinal electric field with no witness beam
        (dashed blue line) and loaded field (whole blue line) along the beam axis. The beam density along the axis for
        both beams are shown in red.
        \textbf{Bottom plot:} plasma densities along the beam axis for a drive beam with no witness beam (dashed green
        line), witness beam with no drive beam (dash-dotted green line), and both beams present (whole green line).
        $\xi = z - tc$ is the position in the simulation box, moving towards the left.}
\end{figure}

\begin{figure}[hbt]
    \includegraphics[width=\linewidth,trim={2mm 0mm 2mm 0mm},clip]{figures/beamPhaseSpace}
    \caption{\label{Fig:BeamPS} Phase space charge distribution of a $100\unit{pC}$, $60\unit{\mu m}$ long witness beam
        after $4\unit{m}$ of plasma. The mean momentum is $1.67\unit{GeV/c}$ with an RMS energy spread
        of $87\unit{MeV/c}$ ($5.2\%$) for the full beam.}
\end{figure}

@@

\begin{figure*}[hbt]
    \includegraphics[width=\linewidth,trim={0mm 0mm 0mm 0mm},clip]{figures/beamFilamentationAll}
    \caption{\label{Fig:BeamFilament} Plots \textbf{a} to \textbf{f} show the transverse phase space of the offset
        electron beam at different plasma positions. Plot \textbf{g} shows the macro particle mean position, and plot
        \textbf{h} their RMS spread. Plots \textbf{a}, \textbf{b} and \textbf{c}, as well as the blue lines in plots
        \textbf{g} and \textbf{h} represent particles not in the ion column (see Fig.~\ref{Fig:BeamEmitt}), with
        position $1.40\unit{\mu m} < \xi < 1.42\unit{\mu m}$. Plots \textbf{d}, \textbf{e} and \textbf{f}, and the red
        lines in plots \textbf{g} and \textbf{h} represent particles in the ion column with position
        $1.55\unit{\mu m} < \xi < 1.57\unit{\mu m}$.}
\end{figure*}

@@

Figure \ref{Fig:BeamLoading} shows the results of QuickPIC simulations of the initial time step, for the base case
parameters. The $E_{z}$-field generated by the proton drive beam is seen as the blue line, shown with and without the
electron beam present. With a proton beam density $n_{pb} \simeq n_{0}$, the wakefields are in the quasi-linear regime
\cite{rosenzweig:2010}. The dashed green line in the lower part of figure \ref{Fig:BeamLoading} shows that the on-axis
plasma density has a depletion of $67\%$, close to what we see in full scale reference simulations for AWAKE Run 2
\cite{awake_collaboration:2016}.

@@

The witness beam generates its own wakefield which loads the $E_{z}$ field generated by the drive beam. With an ideally
shaped electron beam charge profile it is possible to optimally load the field in such a way that the accelerating field
is constant along the beam \cite{katsouleas:1987, tzoufras:2009}. Gaussian beams, as assumed in these studies, cannot
completely flatten the electric field in the tails of the charge distribution, and our base case beam therefore has a
tail in energy both at the front and the back of the beam, as illustrated in figure \ref{Fig:BeamPS}. The bulk of the
beam, however, sees a relatively flat field.
% EA: I rewrote this to the sentence after the comments
%we can acceleratoreHowever, with such a narrow beam when matched beam to the plasma density (equation
%\ref{EQ:MatchedB}), we are limited on how much total charge we can accelerate without reaching charge densities that
%will overload the field.
%  EA: shall we say the below?  The sentence asks for more detailed studies of the dependence of the loading of the transverse fields.
%The amounts of charge required to load the field depends both on the witness bunch length as well as the transverse size.

The electron beam density is $n_{eb} \approx 35\times n_{0}$. This means that the witness beam's own wakefield is in the
fully non-linear regime, where the space charge force is sufficient to blow-out all plasma electrons resulting in the
formation of an ion column (see figure \ref{Fig:BeamLoading}, bottom). This ion column, as is well known
\cite{rosenzweig:1991}, provides a linear focusing force on the part the electron beam within the column, and therefore
prevents emittance growth for this part of the beam.  This effect is shown for our base case in figure
\ref{Fig:PlasmaDenTWake}. The focusing field has a gradient of $20\unit{kT/m}$ near the beam axis, corresponding to the
matched field gradient.

@@

%~ The witness beam's self-generated bubble prevents the tailing end of the beam to see any significant emittance growth.
% EA: I cut the line below, we are still discussing the base case
%For the majority of the cases we studied, that maintained a stable accelerating structure, 
 %PMDue to the ion column focusing, 
Simulation results show that about $70-80\%$ of the electron beam longitudinal slices retain their initial emittance
after the propagation in the plasma. Figure \ref{Fig:BeamEmitt} shows the slice emittance along the beam for the base
case, sampled after propagating through $0$, $4$, $40$ and $100\unit{m}$ of plasma. Emittance growth mainly occurs in
first few meters, and no significant emittance growth is observed after this for propagation lengths up to
$100\unit{m}$. The head of the beam does not benefit from the full ion column focusing, but since the proton beam
creates a quasi-linear wake, the emittance of the head of the beam still stabilizes after some time. For the
$100\unit{m}$ simulation, the drive beam energy was increased to $7\unit{TeV}$ (LHC energy) to prevent de-phasing, as
de-phasing starts to become a significant effect for the SPS beam of $400\unit{GeV}$ after about $50\unit{m}$.
